---
title: "Final 250a project!"
author: "Yael Reuben Ben"
date: "May 27, 2016"
output: pdf_document
---

Connecting Rstudio to out H20 cluster

```{r}
library(h2o)
library(dplyr)
library(ggplot2)
library(tidyr)

cluster <- h2o.init(ip="52.53.176.50", strict_version_check = FALSE)
h2o.clusterStatus()
```

Read data frame - that we loaded to Flow

```{r, echo=FALSE}
orig <- h2o.getFrame("X1929.hex")
summary(orig)
```

Looking at global wearming

## Average temp increases over the years

```{r}
## Plotting avg
temp_by_year <- h2o.group_by(orig, by="Year",
                           nrow("temp"), sum("temp"), mean("temp"), sd("temp"))
# summary(temp_by_year)
# head(temp_by_year)

df <- as.data.frame(temp_by_year)
df %>%
  ggplot() +
    geom_line(mapping = aes(x = Year, y = mean_temp))

# incomplete data for 2016
df %>%
  filter(Year != 2016) %>%
  ggplot() +
    geom_line(mapping = aes(x = Year, y = mean_temp))
```


Adding station data - joining

```{r}
# stations <- h2o.getFrame("emshr_lite.hex")
# h2o.summary(stations)
# head(stations)
# orig$WBAN <- as.factor(orig$WBAN)
# 
# j <- h2o.merge(orig, stations)
# j_allx <- h2o.merge(orig, stations, all.x = TRUE)
# 
# nrow(orig)
# nrow(j)
# nrow(j_allx)
# 
# h2o.summary(j_allx)

```

In the first try we saw the data set wasn't parsing correctly. 
We found another data set and tried again:
```{r}
# try again new data. forst one has a parsing problem becuase of fixed width
stations2 <- h2o.getFrame("Key_Frame__weather_stations.hex")
j2 <- h2o.merge(orig, stations2)
nrow(j2)
names(j2)
h2o.unique(j2$stateprov)

# Looks better!
```

Plotting global wearming in specific location

Alaska colder than Texas: trends per-station

```{r}
j2.TX_AK <- j2[j2$stateprov == "TX" | j2$stateprov == "AK",] 
# head(j2.TX_AK)
# h2o.summary(j2.TX_AK)

j2.TX_AK.df <- as.data.frame(j2.TX_AK)

j2.TX_AK.df %>%
  filter(Station != 999999, Year != 2016) %>%
  ggplot(mapping = aes(x = Year, y = temp, color = as.factor(stateprov))) +
    geom_point() +
    geom_smooth()

```

Winter colder than summer: trends per-day-of-year
```{r}
names(orig)
summary(orig)
temp_by_md <- h2o.group_by(orig, by="MonthDay",
                           nrow("temp"), sum("temp"), mean("temp"), sd("temp"))
temp_by_md.df <- as.data.frame(temp_by_md)

temp_by_md.df  %>%
  ggplot() +
    geom_line(mapping = aes(x = MonthDay, y = mean_temp))

###
```


Find increases in weather extremes
Increases in tornadoes

```{r}
h2o.summary(orig)
tornado_by_year <- h2o.group_by(orig, by="Year",
                           nrow("tornado"), sum("tornado"), 
                           mean("tornado"), sd("tornado"))

tornado_by_year.df <- as.data.frame(tornado_by_year)

tornado_by_year.df %>%
  filter(Year != 2016) %>%
  ggplot() +
    geom_line(mapping = aes(x = Year, y = sum_tornado))
```

Increases in std-dev of min & max temps
```{r}

temp_by_year <- h2o.group_by(orig, by="Year",
                           max("temp"), min("temp"), 
                           mean("temp"), sd("temp"))

temp_by_year.df <- as.data.frame(temp_by_year)

# temp
temp_by_year.df %>%
  filter(Year != 2016) %>%
  ggplot() +
    geom_line(mapping = aes(x = Year, y = max_temp, color = "max_temp")) +
    geom_line(mapping = aes(x = Year, y = min_temp, color = "min_temp")) 

#sd temp
temp_by_year.df %>%
  filter(Year != 2016) %>%
  ggplot() +
    geom_line(mapping = aes(x = Year, y = sdev_temp))

```


Predict temp for given day & place

step1 - created basic model in flow
  a. split data 25 - 75
  b. pick cols

```{r}
h2o.getModel("glm-da2b9e9f-1d54-4459-8521-fb5119f3f434")
```

We see that snow and thunder are strong predictors of temp and R2 :  0.3234671
BUT this is not useful becuase you only know them at the moment
thus it is not really a prediction


Now we are going to try only station and date in order to see if we can 
predict using this
```{r}
h2o.getModel("glm-0a2dfcb7-f79b-4c62-8db1-d5b8b4d31ffb")
```
It works very poorly - r2 = 0.08

We noticed that ittreads station and WBAN as numeric. we converted to enum and ran again:

```{r}
h2o.getModel("glm-f1e8f8e6-3dee-4626-ac7c-0b074df6a606")
```

got worse - r2 = 0.02

In order to improve the baseline model we are trying to introduce more features:

Using location data from the station df

```{r}
# j2.split <- h2o.splitFrame(j2)
# j2.split.75 <- j2.split[[1]]  
# j2.split.25 <- j2.split[[2]]  
# 

j2.glm_3 <- h2o.glm(x = c("Station", "elevation", "MonthDay",
                        "Year", "climate_division"), y = "temp", 
                  training_frame = j2, nfolds = 3, family = "gaussian")
j2.glm_3@model$cross_validation_metrics
```

R2 = 0.11 Boo

Lets add at lng information from out enriched data set

```{r}
j2.glm_3_lat <- h2o.glm(x = c("Station", "elevation", "latitude", "longitude",
                              "MonthDay","Year", "climate_division"), y = "temp", 
                  training_frame = j2, nfolds = 3, family = "gaussian")
j2.glm_3_lat@model$cross_validation_metrics
summary(j2.glm_3)
```

It got better! R2 = 0.29.

The lat long matter! wohoo

Now how about a different algorithm

```{r}
j2.rf <- h2o.randomForest(x = c("Station", "elevation", "latitude", "longitude",
                              "MonthDay","Year", "climate_division"), y = "temp", 
                  max_depth = 5, ntrees = 10, training_frame = j2, nfolds = 3)
j2.rf@model$cross_validation_metrics
summary(j2.rf)
```

WOW we won R2 = 0.5577518

Random forest for the win!
